Finite Complete Suites for CSP Refinement Testing

### 2018-11-29 scp-journal/article/complexity.tex
Finished revision

Since we have finite complete CSP test suites, it is useful for the first
time to calculate how many test executions are needed when using them.
Previous work did not consider sufficient conditions for finiteness, so
complexity was not a concern. We answer the following questions. (1)~What is
the worst-case bound on the number of test executions to be performed to
verify an SUT with respect to failures refinement, when we use our test
suite? (2)~What is the worst-case bound for trace refinement? (3)~Is it
possible to reduce the maximal length of traces when testing for failures or
trace refinement? We consider the first question~(1) in
Section~\ref{section:complexity:failures}), where we also discuss whether it
it is possible to reduce the number of test executions with a different test
suite.  With the answer to question~(1), question~(2) is a fairly simple
consequence we discuss briefly also in
Section~\ref{section:complexity:failures}. Question~(3) is the subject of
Section~\ref{section:complexity:length}.

@@

% -------------------------------------------------------------------------
\paragraph{Estimate for the Maximal Number of Trace Test Executions}
%\label{section:complexity:traces}
According to Theorem~\ref{th:tracetest}, a complete test suite checking trace
refinement just contains the adaptive test case $U_T(pq-1)$. As derived for
$U_F(j)$ above, the maximal number of executions to be performed by
$(Q\parallel[\Sigma] U_T(pq-1))$ is of order
$O\big(\card{\Sigma}^{pq-1}\big)$.

@@

\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.
%
\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & (k < p-1) \& \big( (a \then P(k)) \intchoice ( b \then P(k+1))\big)
\\ & & \extchoice
\\ & & (k = p-1) \& (a \then P(k))
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big)
\\ & & \extchoice
\\ & & ( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}
%
The normalised transition graphs of $P$ and $Q$ are depicted in
Fig.~\ref{fig:examplepq} for the case $p=3,\ q=4$. Using FDR4, it can be
shown for concrete values of $p$ and $q$ that the  ``test passed conditions''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\quad
\mathrm{and}
\quad
(\epass\then\Stop) \lessdet_T (Q\parallel[\Sigma] U_T(j))\hide \Sigma
\]
hold for $j = 0,\dots,pq-2$. So, none of the test cases $U_F(j)$ and $U_T(j)$
are capable of detecting failures and trace refinement violations, if they
only check traces up to length $pq-1$. (We recall that this corresponds to
$j\le pq-2$).

$Q$, however, neither conforms to $P$ in the failures refinement relation,
nor in the trace refinement relation. This can only be seen when executing
the test $U_F(pq-1)$ and $U_T(pq-1)$, respectively. These tests fail, so this
shows that $P\not\lessdet_F Q$ and $P\not\lessdet_T Q$ according to
Theorem~\ref{th:failurestest} and Theorem~\ref{th:tracetest}. Moreover, this
shows that the maximal trace length $pq$ to be investigated in the tests
cannot be further reduced without losing the completeness property of the
test suites. \xbox
\end{example}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any $p,q \geq 2$,
%\in\mathbb{N}$,
there exist reference processes $P$ with $p$ states and
implementation processes $Q$ with $q$ states, such that a violation of the
trace refinement property can only be detected with a trace of length $pq$.
This is proven in the following
theorem. %In the proof, we use the processes $P$ and $Q$ introduced in Example~\ref{ex:pq}.

@@

% ----------------------------------------------------------------------------
\begin{theorem}\label{th:maxtracelen}
Let $2\le p,q \in\mathbb{N}$. Then there exists a reference process $P$ and an
implementation process $Q$ with the following properties.
\begin{enumerate}
\item $G(P)$ has $p$ states.
\item $G(Q)$ has $q$ states.
\item $P\not\lessdet_T Q$, and therefore, also $P\not\lessdet_F Q$.
\item $\forall s\in\trc(Q): \#s < pq\implies s\in\trc(P)$.
\item $Q\ conf\ P$.
\end{enumerate}
As a consequence, the upper bound $pq$ for the length of traces to be tested when checking for failures refinement or trace refinement
cannot be reduced without losing the test suite's completeness property.
\end{theorem}
% ----------------------------------------------------------------------------
\begin{proof}
Given $2\le p,q \in\mathbb{N}$, define reference process $P$ and
implementation process $Q$ as in Example~\ref{ex:pq}. It is trivial to see
that $G(P)$ has $p$ nodes and $G(Q)$ has $q$ nodes, so statements 1 and 2 of
the theorem hold.

@@

To prove statement~5, we observe that the specifications of $P$ and $Q$ lead
to the following sets of minimal acceptances.  In these definition, the
expression $(s\cnt b)$ denotes the number of $b$ events occurring in trace
$s$.
\[
\minaccs(P/s) = \left\{
\begin{array}{ll}
\{ \{a\}, \{b\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) <p-1$.}
\\
\{ \{a\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) = p-1$.}
\end{array}
\right.
\also
\minaccs(Q/s) = \left\{
\begin{array}{ll}
\{ \{a\}  \} & \text{for all $s\in\trc(Q)$ with $\#s \neq 0\mod(q-1)$.}
\\
\{ \{a,b\} \} & \text{for all $s\in\trc(P)$ with $\#s = 0\mod(q-1)$.}
\end{array}
\right.
\]
So, the minimal acceptance set $A_P = \{a\}$ contained in every
$\minaccs(P/s)$ fulfils $A_P \subseteq A_Q$ for any $A_Q\in\minaccs(Q/s)$,
when $s\in\trc(P)\cap \trc(Q)$. Now Lemma~\ref{lemma:tgtrcref}, in particular
(\ref{eq:failrefb}), can be applied to conclude that $Q\ conf\ P$. \xbox
\end{proof}
%
Since Theorem~\ref{th:maxtracelen} just states that a violation of trace
refinement may remain undetected if only traces shorter than $pq$ are checked
during tests, it can also be applied to our trace-refinement tests.
Therefore, test suites $\{ U_T(j) \}$ with $j<pq-1$ are not complete. It is
discussed next in Section~\ref{sec:conc} how the number of test traces to be
executed by complete test suites for failures or trace refinement can still
be reduced {\it without} reducing the maximal length.

>>>>

Since we have finite complete CSP test suites, it is useful for the first
time to calculate how many test executions are needed when using them.
Previous work did not consider sufficient conditions for finiteness, so
complexity was not a concern. We answer the following questions. (1)~What is
the worst-case bound on the number of test executions to be performed to
verify an SUT with respect to failures refinement, when we use our test
suite? (2)~What is the worst-case bound for traces refinement? (3)~Is it
possible to reduce the maximal length of traces when testing for failures or
traces refinement? We consider the first question~(1) in
Section~\ref{section:complexity:failures}), where we also discuss whether it
it is possible to reduce the number of test executions with a different test
suite.  With the answer to question~(1), question~(2) is a fairly simple
consequence we discuss briefly also in
Section~\ref{section:complexity:failures}. Question~(3) is the subject of
Section~\ref{section:complexity:length}.

@@

% -------------------------------------------------------------------------
\paragraph{Estimate for the Maximal Number of Trace Test Executions}
%\label{section:complexity:traces}
According to Theorem~\ref{th:tracetest}, a complete test suite checking
traces refinement just contains the adaptive test case $U_T(pq-1)$. As
derived for $U_F(j)$ above, the maximal number of executions to be performed
by $(Q\parallel[\Sigma] U_T(pq-1))$ is of order
$O\big(\card{\Sigma}^{pq-1}\big)$.

@@

\begin{example}\label{ex:pq}
Consider the CSP reference process $P$ and an erroneous implementation $Q$
specified as follows.
%
\begin{eqnarray*}
P & = &  P(0)
\\
P(k) & = & (k < p-1) \& \big( (a \then P(k)) \intchoice ( b \then P(k+1))\big)
\\ & & \extchoice
\\ & & (k = p-1) \& (a \then P(k))
\\
Q & = & Q(0)
\\
Q(k) & = & (k < q-1) \& \big( a \then Q(k+1)    \big)
\\ & & \extchoice
\\ & & ( k = q-1)\& \big( a\then Q(0) \extchoice b\then Q(0)  \big)
\end{eqnarray*}
%
The normalised transition graphs of $P$ and $Q$ are depicted in
Fig.~\ref{fig:examplepq} for the case $p=3,\ q=4$. Using FDR4, it can be
shown for concrete values of $p$ and $q$ that the  ``test passed conditions''
\[
(\epass\then\Stop) \lessdet_F (Q\parallel[\Sigma] U_F(j))\hide \Sigma
\quad
\mathrm{and}
\quad
(\epass\then\Stop) \lessdet_T (Q\parallel[\Sigma] U_T(j))\hide \Sigma
\]
hold for $j = 0,\dots,pq-2$. So, none of the test cases $U_F(j)$ and $U_T(j)$
are capable of detecting failures and traces-refinement violations, if they
only check traces up to length $pq-1$. (We recall that this corresponds to
$j\le pq-2$).

$Q$, however, neither conforms to $P$ in the failures refinement relation,
nor in the traces-refinement relation. This can only be seen when executing
the test $U_F(pq-1)$ and $U_T(pq-1)$, respectively. These tests fail, so this
shows that $P\not\lessdet_F Q$ and $P\not\lessdet_T Q$ according to
Theorem~\ref{th:failurestest} and Theorem~\ref{th:tracetest}. Moreover, this
shows that the maximal trace length $pq$ to be investigated in the tests
cannot be further reduced without losing the completeness property of the
test suites. \xbox
\end{example}
%
Generalising Example~\ref{ex:pq}, it can be shown that for any $p,q \geq 2$,
%\in\mathbb{N}$,
there exist reference processes $P$ with $p$ states and implementation
processes $Q$ with $q$ states, such that a violation of the traces-refinement
property can only be detected with a trace of length $pq$. This is proven in
the following
theorem. %In the proof, we use the processes $P$ and $Q$ introduced in Example~\ref{ex:pq}.

@@

% ----------------------------------------------------------------------------
\begin{theorem}\label{th:maxtracelen}
Let $2\le p,q \in\mathbb{N}$. Then there exists a reference process $P$ and an
implementation process $Q$ with the following properties.
\begin{enumerate}
\item $G(P)$ has $p$ states.
\item $G(Q)$ has $q$ states.
\item $P\not\lessdet_T Q$, and therefore, also $P\not\lessdet_F Q$.
\item $\forall s\in\trc(Q): \#s < pq\implies s\in\trc(P)$.
\item $Q\ conf\ P$.
\end{enumerate}
As a consequence, the upper bound $pq$ for the length of traces to be tested
when checking for failures refinement or traces refinement cannot be reduced
without losing the test suite's completeness property.
\end{theorem}
% ----------------------------------------------------------------------------
\begin{proof}
Given $2\le p,q \in\mathbb{N}$, define reference process $P$ and
implementation process $Q$ as in Example~\ref{ex:pq}. It is trivial to see
that $G(P)$ has $p$ nodes and $G(Q)$ has $q$ nodes, so statements 1 and 2 of
the theorem hold.

@@

To prove statement~5, we observe that the specifications of $P$ and $Q$ lead
to the following sets of minimal acceptances.  In these definition, the
expression $(s\cnt b)$ denotes the number of $b$ events occurring in trace
$s$.
\[
\minaccs(P/s) = \left\{
\begin{array}{ll}
\{ \{a\}, \{b\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) <p-1$.}
\\
\{ \{a\} \} & \text{for all $s\in\trc(P)$ with $(s\cnt b) = p-1$.}
\end{array}
\right.
\also
\minaccs(Q/s) = \left\{
\begin{array}{ll}
\{ \{a\}  \} & \text{for all $s\in\trc(Q)$ with $\#s \neq 0\mod(q-1)$.}
\\
\{ \{a,b\} \} & \text{for all $s\in\trc(P)$ with $\#s = 0\mod(q-1)$.}
\end{array}
\right.
\]
So, the minimal acceptance set $A_P = \{a\}$ contained in every
$\minaccs(P/s)$ fulfils $A_P \subseteq A_Q$ for any $A_Q\in\minaccs(Q/s)$,
when $s\in\trc(P)\cap \trc(Q)$. Now Lemma~\ref{lemma:tgtrcref}, in particular
(\ref{eq:failrefb}), can be applied to conclude that $Q\ conf\ P$. \xbox
\end{proof}
%
It is discussed next in Section~\ref{sec:conc} how the number of test traces
to be executed by complete test suites for failures or traces refinement can
still be reduced {\it without} reducing the maximal length.

### 2018-11-29 scp-journal/article/finitecomplete-failures.tex
Finished revision

It is necessary to use failures refinement in the definition above, and not
just trace refinement:~$(Q\parallel[\Sigma] U_F(j)) \hide \Sigma$ may have
the same visible traces $\varepsilon$ and $\epass$ as the ``Test Passed
Process'' $(\epass\then\Stop)$. However, the former may nondeterministically
refuse $\epass$, due to a deadlock occurring when a faulty SUT process
executes concurrently with $U_F(j,k,n)$ executing branch~(\ref{eq:ufd}), when
the guard condition $(k = j\wedge \minhits(n)\neq\varnothing)$ evaluates to
$\ist$. This is explained further in the next paragraphs. Alternatively, a
faulty SUT $Q$ might internally deadlock after a trace $s$ where $\#s < j$
and $\minhits(G(P)/s) \neq \varnothing$, so that the process
$(Q\parallel[\Sigma] U_F(j))/s$ deadlocks as well.

>>>>

It is necessary to use failures refinement in the definition above, and not
just traces refinement:~$(Q\parallel[\Sigma] U_F(j)) \hide \Sigma$ may have
the same visible traces $\varepsilon$ and $\epass$ as the ``Test Passed
Process'' $(\epass\then\Stop)$. However, the former may nondeterministically
refuse $\epass$, due to a deadlock occurring when a faulty SUT process
executes concurrently with $U_F(j,k,n)$ executing branch~(\ref{eq:ufd}), when
the guard condition $(k = j\wedge \minhits(n)\neq\varnothing)$ evaluates to
$\ist$. This is explained further in the next paragraphs. Alternatively, a
faulty SUT $Q$ might internally deadlock after a trace $s$ where $\#s < j$
and $\minhits(G(P)/s) \neq \varnothing$, so that the process
$(Q\parallel[\Sigma] U_F(j))/s$ deadlocks as well.

### 2018-11-29 scp-journal/article/finitecomplete-trace.tex
Finished revision

% ==========================================================================
\section{Finite Complete Test Suites for CSP Trace Refinement}
\label{sec:finitecomplete}
% ==========================================================================

For establishing trace refinement, the following class of adaptive test cases
are used for a given reference process $P$ and  integers $j \ge 0$. Just as
for the tests developed  in Section~\ref{sec:finitecompletefails} to verify
failures refinement, the tests for trace refinement are derived from the
reference model's transition graph
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)).
$$
In contrast to the tests for failures refinement~(\ref{eq:UFP}), however, we
do not need to check the SUT with respect to its acceptance of hitting sets.
Therefore, these do not occur in the specification of the test cases below.
We use the condition on acceptances $\minaccs(n) = \{ \varnothing \}$ instead
of the condition on hitting sets $\minhits(n) = \varnothing$ in branch
(\ref{eq:utb}). From (\ref{eq:minhitminaccempty}) we know that these
conditions are equivalent, but, with the use of $\minaccs(n) = \{ \varnothing
\}$, \pagebreak we make it unnecessary to calculate hitting sets for
generating these tests from $G(P)$, which is expensive.

@@

The existence of complete, finite test suites is expressed in analogy to
Theorem~\ref{th:failurestest}. A noteworthy difference is that the complete
suite for trace refinement just needs the single adaptive test case
$U_T(pq-1)$, while failures refinement requires the execution of $\{
U_F(0),\dots,U_F(pq-1)\}$. The reason is that $U_T(pq-1)$ identifies trace
errors for all traces up to length $pq$, while $U_F(pq-1)$ only probes for
erroneous acceptances at the end of each trace of length $(pq -1)$.
%
% -------------------------------------------------------------------------
\begin{theorem}\label{th:tracetest}
Let $P$ be a non-terminating, divergence-free CSP process over alphabet $\Sigma$ whose
normalised transition graph $G(P)$ has $p$ states. Define fault domain ${\cal
D}$ as the set of all non-terminating, divergence-free CSP processes over alphabet $\Sigma$,
whose transition graph has at most $q$ states with $q \ge p$. Then the test
suite
\[
\TS_T = \{ U_T(pq-1)   \}
\]
is complete with respect to ${\cal F} = (P,\lessdet_T,{\cal D})$.
\xbox
\end{theorem}
% -------------------------------------------------------------------------
%\begin{proof}
%The theorem follows directly
%from Step~1 in the proof of Lemma~\ref{lemma:mainfsound} and
%Case~1 in the proof of Lemma~\ref{lemma:mainfexhaustive}.
%\xbox
%\end{proof}
%%
%Examples are provided after our discussion of size of the test suites.
%
As for Theorem~\ref{th:failurestest}, the proof is structured in two lemmas,
the first ensuring soundness, and the second exhaustiveness.

>>>>

% ==========================================================================
\section{Finite Complete Test Suites for CSP Traces Refinement}
\label{sec:finitecomplete}
% ==========================================================================

For establishing traces refinement, the following class of adaptive test
cases are used for a given reference process $P$ and  integers $j \ge 0$.
Just as for the tests developed  in Section~\ref{sec:finitecompletefails} to
verify failures refinement, the tests for traces refinement are derived from
the reference model's transition graph
$$
G(P) = ( N, \ii n, \Sigma, t : N\times\Sigma \pfun N, r : N \fun \mathbb{P}\mathbb{P}(\Sigma)).
$$
In contrast to the tests for failures refinement~(\ref{eq:UFP}), however, we
do not need to check the SUT with respect to its acceptance of hitting sets.
Therefore, these do not occur in the specification of the test cases below.
We use the condition on acceptances $\minaccs(n) = \{ \varnothing \}$ instead
of the condition on hitting sets $\minhits(n) = \varnothing$ in branch
(\ref{eq:utb}). From (\ref{eq:minhitminaccempty}) we know that these
conditions are equivalent, but, with the use of $\minaccs(n) = \{ \varnothing
\}$, \pagebreak we make it unnecessary to calculate hitting sets for
generating these tests from $G(P)$, which is expensive.

@@

The existence of complete, finite test suites is expressed in analogy to
Theorem~\ref{th:failurestest}. A noteworthy difference is that the complete
suite for traces refinement just needs the single adaptive test case
$U_T(pq-1)$, while failures refinement requires the execution of $\{
U_F(0),\dots,U_F(pq-1)\}$. The reason is that $U_T(pq-1)$ identifies trace
errors for all traces up to length $pq$, while $U_F(pq-1)$ only probes for
erroneous acceptances at the end of each trace of length $(pq -1)$.
%
% -------------------------------------------------------------------------
\begin{theorem}\label{th:tracetest}
Let $P$ be a non-terminating, divergence-free CSP process over alphabet $\Sigma$ whose
normalised transition graph $G(P)$ has $p$ states. Define fault domain ${\cal
D}$ as the set of all non-terminating, divergence-free CSP processes over alphabet $\Sigma$,
whose transition graph has at most $q$ states with $q \ge p$. Then the test
suite
\[
\TS_T = \{ U_T(pq-1)   \}
\]
is complete with respect to ${\cal F} = (P,\lessdet_T,{\cal D})$.
\xbox
\end{theorem}
% -------------------------------------------------------------------------
%\begin{proof}
%The theorem follows directly
%from Step~1 in the proof of Lemma~\ref{lemma:mainfsound} and
%Case~1 in the proof of Lemma~\ref{lemma:mainfexhaustive}.
%\xbox
%\end{proof}
%%
%Examples are provided after our discussion of size of the test suites.
%
As for Theorem~\ref{th:failurestest}, the proof is structured in two lemmas,
the first ensuring soundness, and the second exhaustiveness.
